Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2019-10-31 ASA2013/asa2013.tex
2019-08-18
2019-10-30

\documentclass[
a4paper,10pt
]{ica2013_2}

@@

\newcommand\1{\vec 1}
\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

@@

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
\usepackage{glossaries}
\usepackage{common/slashbox}

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
\end{figure}
The basic idea behind the MVDR method is to compute the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signals coming from the pixel of interest. When the data is delayed this can be formulated as follows:
\begin{gather}
\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
This is a convex optimization problem with the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

@@

\bibliographystyle{jasanum}
% \bibliography{../../Library/library}
\bibliography{library}

>>>>

\documentclass[
a4paper,10pt
]{common/ica2013_2}

@@

\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

@@

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
% \usepackage{glossaries}

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
\end{figure}
The basic idea behind the MVDR method is to compute the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signals coming from the pixel of interest. When the data is delayed this can be formulated as follows:
\begin{gather}
\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
This is a convex optimization problem with the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

@@

\bibliographystyle{jasanum}
% \bibliography{../../Library/library}
% \bibliography{library}

### 2019-10-31 ECUA2012/gpu_capon_ecua2012.tex
2019-10-30

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$. To simplify the upcoming notation, let sample $x_m[n]$ delayed by $\Delta_m[n]$ be written as:
\begin{align}
x^\Delta_m = x_m[n-\Delta_m[n]].
\end{align}
The beamformer output $z[n]$ is then simply the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x^\Delta_0[n]\\x^\Delta_1[n]\\\vdots\\x^\Delta_{M-1}[n]}.\label{z}
\end{align}
If the weights in (\ref{z}) were static, this would be referred to as the \gls{DAS} beamformer. The \gls{MV} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x[n]\H\}$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by computing the sample covariance matrix, $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009a}.

>>>>

Consider an $M$ element phased array. Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$. To simplify the upcoming notation, let sample $x_m[n]$ delayed by $\Delta_m[n]$ be written as:
\begin{align}
x^\Delta_m = x_m[n-\Delta_m[n]].
\end{align}
The beamformer output $z[n]$ is then simply the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x^\Delta_0[n]\\x^\Delta_1[n]\\\vdots\\x^\Delta_{M-1}[n]}.\label{z}
\end{align}
If the weights in (\ref{z}) were static, this would be referred to as the \gls{DAS} beamformer. The \gls{MV} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction. The solution to this optimization problem is given as:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x[n]\H\}$ is the spatial covariance matrix. Since $\R$ is unknown, it must be estimated. We will do this by computing the sample covariance matrix, $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009}.

### 2019-10-31 HRUIofMovingObjects/jrnl_article.tex
2019-10-30

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). Dotted vertical lines indicate position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse. Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point. The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first. The beam density was calculated  based on (\ref{eq:resolution}) with a speed of sound equal to 1540. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

>>>>

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). Dotted vertical lines indicate position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse. Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point. The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first. The beam density was calculated  based on (\ref{eq:resolution}) with a speed of sound equal to 1540. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

### 2019-10-31 HRUIofMovingObjects/jrnl_article_1st.tex
2019-10-30

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). Dotted vertical lines indicate position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse. The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

>>>>

Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}). Dotted vertical lines indicate position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and 1.5 cycles long pulse. The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point. We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE). We therefore conclude that the system is locally laterally shift-invariant. %In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.

### 2019-10-31 HRUIonGPU/1st submit/jrnl_article.tex
2019-10-30

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of temporal samples included, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness: 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig. \ref{fig:phantom}). The simulation was performed using a 64-element, $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

>>>>

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of temporal samples included, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness: 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.

@@

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krim1996}. Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Fig. \ref{fig:phantom}). The simulation was performed using a 64-element, $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

### 2019-10-31 HRUIonGPU/document.tex
2019-06-26
2019-09-01
2019-10-30

% denne må utelates i IUS-artikkelen
In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{Asen2012}. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. In \cite{Asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide. Since the execution time of Capon beamforming is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we extend the work done in \cite{Asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation. The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon when applied in narrow-beam systems.
%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
%
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{II_eq:R}
\end{align}
%
where  $N_K = 2K + 1$ is the number of temporal samples included, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness:
%
\begin{align}\label{II_eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
%
The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
%
\begin{align}\label{II_eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
%
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{II_eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{Asen2012} and \cite{Buskenes}. This implementation was our starting point when the BS-Capon beamformer was implemented.

@@

In \cite{Asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
can be reduced by realizing that calculations in (\ref{II_eq:R}) overlap across subarrays. %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
%\begin{align}
%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
%&- x_{i-1}[n']x_{j-1}[n']^H). \label{II_eq:sliding}
%\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. 
%In that case, 
The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
%The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{II_fig:gpulayout}). 
Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024). %However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

@@

%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig. \ref{II_fig:phantom}). The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is frame $15$ in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

@@

The phantom used contained two point targets with 30 dB amplitude located at 90 mm depth. The two point targets were placed inside speckle, with 0 dB in amplitude, covering an area from $\pm90\degree$ in azimuth. The phantom was simulated using Field II \cite{Jensen1992, Jensen1996a}. The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate of 25 FPS, and the relative speed of the two point scatterers, 3 mm/s. To avoid having to deal with signal cancellation and errors caused by not hitting each point with a receive beam, 16x oversampling has been used on transmit.

@@

\ifPhdDoc
   \printbibliography[title=References,heading=subbibliography]
%    \bibliographysty
%    \bibliography{library.bib}
%    \print
\else

>>>>

% denne må utelates i IUS-artikkelen
In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes2015, Buskenes2013} and cardiac ultrasound imaging \cite{Asen2012}. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. In \cite{Asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide. Since the execution time of Capon beamforming is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we extend the work done in \cite{Asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation. The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon when applied in narrow-beam systems.
%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
%
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{II_eq:R}
\end{align}
%
where  $N_K = 2K + 1$ is the number of temporal samples included, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness:
%
\begin{align}\label{II_eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
%
The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
%
\begin{align}\label{II_eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
%
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.

@@

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{II_eq:R}) is typically averaged over a long sequence of samples\cite{Krim1996}. Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated. The estimated matrix is therefore often found to be both inaccurate and poorly conditioned. Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}. Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.

@@

In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{Asen2012} and \cite{Buskenes2015}. This implementation was our starting point when the BS-Capon beamformer was implemented.

@@

In \cite{Asen2012, Buskenes2015} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
can be reduced by realizing that calculations in (\ref{II_eq:R}) overlap across subarrays. %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
%\begin{align}
%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
%&- x_{i-1}[n']x_{j-1}[n']^H). \label{II_eq:sliding}
%\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. 
%In that case, 
The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
%The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{II_fig:gpulayout}). 
Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024). %However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

@@

%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Fig. \ref{II_fig:phantom}). The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is frame $15$ in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

@@

The phantom used contained two point targets with 30 dB amplitude located at 90 mm depth. The two point targets were placed inside speckle, with 0 dB in amplitude, covering an area from $\pm90\degree$ in azimuth. The phantom was simulated using Field II \cite{Jensen1992, Jensen1996}. The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate of 25 FPS, and the relative speed of the two point scatterers, 3 mm/s. To avoid having to deal with signal cancellation and errors caused by not hitting each point with a receive beam, 16x oversampling has been used on transmit.

@@

\ifPhdDoc
   \printbibliography[title=References,heading=subbibliography]
   \addcontentsline{toc}{section}{References}
%    \bibliographysty
%    \bibliography{library.bib}
%    \print
\else
